Accurate inference of microbial differential abundance from taxonomically-biased microbiome measurements

### 2021-10-19 differential-abundance.Rmd
Make minor edits
Edits include

- use 'widehat' instead of 'hat' over words in equations
- use 'measurement' and 'measured' instead of 'estimate' and 'estimated'
  for individual samples and change between pairs of samples, reserving
  'estimate' for regression over multiple samples
- edits based on Hypothesis comments on version
  6932587bd1ea9c56638c0ebe3ff7885b79b84e8c

How do the measurement errors described in the previous section impact our ability to estimate the changes in microbial abundances across samples or between different host and environmental conditions?
Though there are many ways to quantitatively define such change, here we restrict our attention to inferring the multiplicative or (equivalently) log fold change in proportions, ratios, and cell densities, as such inference is ubiquitous in microbiome DA analysis and have more direct ecological interpretations (via the processes of exponential growth and death) than other DA measures.
These measures of DA are also affected by taxonomic bias in our model in a manner that is straightforward to understand in terms of the multiplicative errors described in the previous section.

@@

Before considering the stereotypical many-sample DA analysis, it is instructive to 
consider simplest analysis of differential abundance: the estimation of fold changes in abundance between a pair of samples.
This case is relevant for understanding common visualizations for comparing abundances across individual samples, such as the ubiquitous proportion bar plot and abundances-through-time trajectories within a single host or environment, and conceptually bridges the single-sample results of the previous section to the many-sample case.

<!-- P: error in FC proportions -->
The composition-dependent effect of bias on fold error in proportions leads to error in fold-change estimates that is proportional to the inverse change in mean efficiency.
From the error in an individual sample \@ref(eq:prop-error), it follows that the estimated fold change in proportion of taxon $i$ from sample $a$ to sample $b$ is
\begin{align}
  (\#eq:prop-fc-error)
% \tag*{Fold change in proportion}
\underbrace{\frac{\hat{\text{prop}}_{i}(b)}{\hat{\text{prop}}_{i}(a)}} _\text{estimated FC}
  &= \frac
    {\text{prop}_{i}(b) \cdot \cancel{\text{efficiency}_{i}} / {\text{mean efficiency}(b)}}
    {\text{prop}_{i}(a) \cdot \cancel{\text{efficiency}_{i}} / {\text{mean efficiency}(a)}}
\\[0.5ex]
  &=
  \underbrace{\frac{\text{prop}_{i}(b)}{\text{prop}_{i}(a)}}_\text{actual FC}
  \cdot
  \underbrace{\left[\frac{\text{mean efficiency}(b)}{\text{mean efficiency}(a)}\right]^{-1}}_\text{fold error}
  .
\end{align}
The sample-independent efficiency factor of the error cancels, but the sample-dependent mean efficiency does not, leaving an error equal to the inverse of the change in the mean efficiency from $s$ to $t$.

The bottom row of Figure \@ref(fig:error-proportions) illustrates how variation in mean efficiency leads to error in the inferred fold changes between a pair of samples.
The mean efficiency decreases by a factor of 2.6 (FC: 0.4X) from Sample 1 to Sample 2.
Consequently, the FC of each species is measured to be 2.6X larger than the true value.
Though the fold error for all species is the same, the implications depend on the actual FC and correspond to three distinct types of error: an increase in magnitude, a decrease in magnitude, and a change in direction (or sign).
We can see each type of error in Figure 2.
For Species 1, which increases and thus moves in the opposite direction of the mean efficiency, we see an increase in magnitude of the estimated FC (actual FC: 2.3X, measured FC: 6.5X).
For Species 2, which decreases and thus moves in the same direction as the mean efficiency but by a larger factor, we see an decrease in magnitude (actual FC: 0.15X, measured FC: 0.44X).
For Species 2, which decreases by a smaller factor than the mean efficiency, we see a change in direction (actual FC: 0.6X, measured FC: 1.8X), such that the species actually appears to increase!

@@

If other error sources remain negligible, than this dichotomy continues to apply to proportion- and ratio-based density estimates.
The inferred FCs in proportion-based density estimates will be incorrect by a factor equal to the inverse fold change in the mean efficiency, creating magnitude and/or directional errors.
Moreover, any species with a constant density will appear to vary inversely with the mean efficiency.
In contrast, fold changes in ratio-based density estimates remain accurate.

@@

DA analysis of many samples across host or environmental conditions can typically be framed as a regression problem, in which we analyze the relationship between a microbial _response variable_, such as log density of some focal species $i$, and one or more _covariates_, such the pH or temperature of the sampled environment or whether the sample is from a healthy or sick person.
A substantial fraction of DA analyses use minor elaborations on the simple linear regression model, which for a response of log species density can be written
\begin{align}
  (\#eq:regression)
  \log \text{density}_i(a) = \alpha + \beta x(a) + \varepsilon_i(a).
\end{align}
Here $x$ is a continuous covariate (e.g. pH) or a binary covariate (e.g. $x=1$ for treated patients and $x=0$ for controls), $\alpha$ and $\beta$ are regression coefficients, and $\varepsilon_i(a)$ is a mean-zero random variable that reflects the residual (unexplained) variation in the response (log density of species $i$).
Our interest is usually in the coefficient $\beta$ (slope or average difference between conditions) that describes how the species' abundance changes with $x$, while the intercept $\alpha$ captures differences in the baseline abundance and---we hope---measurement efficiency among species.
How does taxonomic bias under our measurement model affect estimates of $\beta$ in the simple linear regression?

Consider the case where the response is log density that has been estimated using proportion-based density estimation (Equation \@ref(eq:density-prop-est)) with error-free estimates of the total density.
If the true log density follows the regression Equation \@ref(eq:regression), then it follows from Equation \@ref(eq:density-prop-error) the estimated log density equals 
\begin{align}
  (\#eq:regression-error)
  \log \hat{\text{density}}_i(a)
%  &= \alpha + \beta x(a) + \varepsilon_i(a) + \log \text{efficiency}_i - \log \text{mean efficiency}(a)
  = [\alpha + \log \text{efficiency}_i] + [\beta - \log \text{mean efficiency}(a)] x(a) + \varepsilon_i(a).
\end{align}
This equation shows that the species-specific portion of the error affects the intercept term while the sample-specific portion (log mean efficiency) affects the slope term.
Thus, as in the case of fold changes between two samples (Equation \@ref(eq:prop-fc-error)), it is the variation in the (log) mean efficiency across samples that we must worry about distorting our DA results.

The variation in measurement error created by the log mean efficiency impacts the point estimate and precision of the regression coefficient $\beta$.
Box \@ref(regression-error) mathematically describes this effect for estimation method of Ordinary Least Squares (OLS) and Maximum Likelihood Estimation (MLE) and Figure \@ref(fig:regression-example) illustrates using simulated data.
Variation in the log mean efficiency that is associated with the covariate $x$ creates a systematic error in the estimated slope $\hat \beta$ equal to the negative of the (scaled) covariance of log mean efficiency with $x$.
The absolute error is the same for all species; however, its relative value depends on the magnitude of the covariance of the log mean efficiency with $x$ relative to that of the response (here, $\log \text{density}_{i}$) with $x$ or, equivalently, the relative magnitudes of their slopes.
As in the case of fold changes between pairs of samples, the net effect can be decreases in magnitude (Species 9, 10, and 1 in Figure \@ref(fig:regression-example)), changes in sign (Species 5), or increases in magnitude (remaining species) depending on these relative values.
Variation in the log mean efficiency that is uncorrelated with $x$ does not systematically distort $\hat \beta$ but does affect its precision, typically leading to increased standard errors as the variation in log mean efficiency effectively acts as an additional source of noise in measured abundance (Figure \@ref(fig:regression-example) D).
The exception is for species whose residual variation is strongly positively correlated with that of log mean efficiency (here, Species 9), which can appear to have less random variation and receive standard errors that are too small.
Decreased magnitudes and increased standard errors can both cause associations to be missed that would otherwise have been detected (Species 10 and 1), while increased magnitudes can turn weak or statistically insignificant associations into strong and statistically significant ones (Species 7, 6 and 4).

@@

```{r regression-example, fig.cap = '(ref:regression-example)'}
svg_path <- here::here(
  "notebook/_posts/2021-08-03-simulate-regression-example/",
  "simulate-regression-example_files/figure-html5/",
  "main-figure-1.svg"
)
if (knitr::is_html_output()) {
  knitr::include_graphics(svg_path)
} else if (knitr::is_latex_output()) {
  cowplot::ggdraw() + cowplot::draw_image(svg_path)
}
rm(svg_path)
```

>>>>

How do the measurement errors described in the previous section impact our ability to estimate the changes in microbial abundances across samples or between different host and environmental conditions?
Though there are many ways to quantitatively define such change, here we restrict our attention to inferring the multiplicative or (equivalently) log fold change in proportions, ratios, and cell densities, which have direct ecological interpretations (via the processes of exponential growth and death) and are ubiquitous in microbiome DA analysis.

@@

Before considering the stereotypical many-sample DA analysis, it is instructive to 
consider simplest analysis of differential abundance: the measurement of fold changes in abundance between a pair of samples.
This case is relevant for understanding common visualizations for comparing abundances across individual samples, such as the ubiquitous proportion bar plot and abundances-through-time trajectories within a single host or environment, and conceptually bridges the single-sample results of the previous section to the many-sample case.

<!-- P: error in FC proportions -->
The composition-dependent effect of bias on fold error in proportions leads to error in fold-change measurements that is proportional to the inverse change in mean efficiency.
From the error in an individual sample (Equation \@ref(eq:prop-error)), it follows that the measured fold change in proportion of taxon $i$ from sample $a$ to sample $b$ is
\begin{align}
  (\#eq:prop-fc-error)
% \tag*{Fold change in proportion}
\underbrace{\frac{\widehat{\text{prop}}_{i}(b)}{\widehat{\text{prop}}_{i}(a)}} _\text{measured FC}
  &= \frac
    {\text{prop}_{i}(b) \cdot \cancel{\text{efficiency}_{i}} / {\text{mean efficiency}(b)}}
    {\text{prop}_{i}(a) \cdot \cancel{\text{efficiency}_{i}} / {\text{mean efficiency}(a)}}
\\[0.5ex]
  &=
  \underbrace{\frac{\text{prop}_{i}(b)}{\text{prop}_{i}(a)}}_\text{actual FC}
  \cdot
  \underbrace{\left[\frac{\text{mean efficiency}(b)}{\text{mean efficiency}(a)}\right]^{-1}}_\text{fold error}
  .
\end{align}
The sample-independent efficiency factor of the error cancels, but the sample-dependent mean efficiency does not, leaving an error equal to the inverse of the change in the mean efficiency from $s$ to $t$.

The bottom row of Figure \@ref(fig:error-proportions) illustrates how variation in mean efficiency leads to error in the inferred fold changes between a pair of samples.
The mean efficiency decreases by a factor of 2.6 (FC: 0.4X) from Sample 1 to Sample 2.
Consequently, the FC of each species is measured to be 2.6X larger than the true value.
Though the fold error for all species is the same, the implications depend on the actual FC and correspond to three distinct types of error: an increase in magnitude, a decrease in magnitude, and a change in direction (or sign).
We can see each type of error in Figure 2.
For Species 1, which increases and thus moves in the opposite direction of the mean efficiency, we see an increase in magnitude of the measured FC (actual FC: 2.3X, measured FC: 6.5X).
For Species 2, which decreases and thus moves in the same direction as the mean efficiency but by a larger factor, we see an decrease in magnitude (actual FC: 0.15X, measured FC: 0.44X).
For Species 2, which decreases by a smaller factor than the mean efficiency, we see a change in direction (actual FC: 0.6X, measured FC: 1.8X), such that the species actually appears to increase!

@@

If other error sources remain negligible (or constant across samples), than this dichotomy continues to apply to proportion- and ratio-based density measurements.
The inferred FCs in proportion-based density measurements will be incorrect by a factor equal to the inverse fold change in the mean efficiency, creating magnitude and/or directional errors.
Moreover, any species with a constant density will appear to vary inversely with the mean efficiency.
In contrast, fold changes in ratio-based density measurements will remain accurate.

@@

DA analysis of many samples across host or environmental conditions can typically be framed as a regression problem, in which we analyze the relationship between a microbial _response variable_, such as log density of some focal species $i$, and one or more _covariates_, such the pH or temperature of the sampled environment or whether the sample is from a healthy or sick person.
A large fraction of DA analyses use elaborations on the simple linear regression model, which for a response of log species density can be written
\begin{align}
  (\#eq:regression)
  \log \text{density}_i(a) = \alpha + \beta x(a) + \varepsilon_i(a).
\end{align}
Here $x$ is a continuous covariate (e.g. pH) or a binary covariate (e.g. $x=1$ for treated patients and $x=0$ for controls), $\alpha$ and $\beta$ are regression coefficients, and $\varepsilon_i(a)$ is a mean-zero random variable that reflects the residual (unexplained) variation in the response (log density of species $i$).
Our interest is usually in the coefficient $\beta$ (slope or average difference between conditions) that describes how the species' abundance changes with $x$, while the intercept $\alpha$ captures differences in the baseline abundance and---we hope---measurement efficiency among species.
How does taxonomic bias under our measurement model affect estimates of $\beta$ in the simple linear regression?

Consider the case where the response is log density that has been estimated using proportion-based density estimation (Equation \@ref(eq:density-prop-meas)) with error-free estimates of the total density.
If the true log density follows the regression Equation \@ref(eq:regression), then it follows from Equation \@ref(eq:density-prop-error) the estimated log density equals 
\begin{align}
  (\#eq:regression-error)
  \log \widehat{\text{density}}_i(a)
%  &= \alpha + \beta x(a) + \varepsilon_i(a) + \log \text{efficiency}_i - \log \text{mean efficiency}(a)
  = [\alpha + \log \text{efficiency}_i] + [\beta - \log \text{mean efficiency}(a)] x(a) + \varepsilon_i(a).
\end{align}
This equation shows that the species-specific portion of the error affects the intercept term while the sample-specific portion (log mean efficiency) affects the slope term.
Thus, as in the case of fold changes between two samples (Equation \@ref(eq:prop-fc-error)), it is the variation in the (log) mean efficiency across samples that we must worry about distorting our DA results.

The variation in measurement error created by the log mean efficiency impacts the point estimate and precision of the regression coefficient $\beta$.
Box \@ref(regression-error) mathematically describes this effect for estimation via Ordinary Least Squares (OLS) or Maximum Likelihood Estimation (MLE), which Figure \@ref(fig:regression-example) illustrates using simulated data.
Variation in the log mean efficiency that is associated with the covariate $x$ creates a systematic error in the estimated slope $\hat \beta$ equal to the negative of the (scaled) covariance of log mean efficiency with $x$.
The absolute error is the same for all species; however, its relative value depends on the magnitude of the covariance of the log mean efficiency with $x$ relative to that of the response (here, $\log \text{density}_{i}$) with $x$ or, equivalently, the relative magnitudes of their slopes.
As in the case of fold changes between pairs of samples, the net effect can be decreases in magnitude (Species 9, 10, and 1 in Figure \@ref(fig:regression-example)), changes in sign (Species 5), or increases in magnitude (remaining species) depending on these relative values.
Variation in the log mean efficiency that is uncorrelated with $x$ does not systematically distort $\hat \beta$ but does affect its precision, typically leading to increased standard errors as the variation in log mean efficiency effectively acts as an additional source of noise in measured abundance (Figure \@ref(fig:regression-example) D).
The exception is for species whose residual variation is strongly positively correlated with that of log mean efficiency (here, Species 9), which can appear to have less random variation and receive standard errors that are too small.
Decreased magnitudes and increased standard errors can both cause associations to be missed that would otherwise have been detected (Species 10 and 1), while increased magnitudes can turn weak or statistically insignificant associations into strong and statistically significant ones (Species 7, 6 and 4).

@@

```{r regression-example, fig.cap = '(ref:regression-example)'}
svg_path <- here::here(
  "notebook/_posts/2021-08-03-simulate-regression-example/",
  "simulate-regression-example_files/figure-html5",
  "main-figure-1.svg"
)
if (knitr::is_html_output()) {
  knitr::include_graphics(svg_path)
} else if (knitr::is_latex_output()) {
  # cowplot::ggdraw() + cowplot::draw_image(svg_path)
  pdf_path <- file.path("/tmp", "bookdown-differential-abundance-theory",
    "regression-example.pdf")
  args <- c(
    "--export-area-page", 
    "--export-type=pdf", 
    str_c("--export-filename=", pdf_path),
    svg_path
  )
  system2("inkscape", args = args)
  knitr::include_graphics(pdf_path)
  rm(pdf_path)
}
rm(svg_path)
```

### 2021-10-19 introduction.Rmd
Make minor edits
Edits include

- use 'widehat' instead of 'hat' over words in equations
- use 'measurement' and 'measured' instead of 'estimate' and 'estimated'
  for individual samples and change between pairs of samples, reserving
  'estimate' for regression over multiple samples
- edits based on Hypothesis comments on version
  6932587bd1ea9c56638c0ebe3ff7885b79b84e8c

<!-- First paragraph: DA analysis -->

The most basic question we can ask about microbial communities after which taxa are found in an ecosystem is, how do the abundances of these taxa vary in their abundance—across space, time, and host or environmental conditions?
Advances in sequencing technology allow us to now simultaneously measure the abundances of 100s to 1000s of species using marker-gene and shotgun-metagenomic sequencing (jointly, MGS).
Although standard MGS measurements lose information about total microbial density—and so are typically used to analyze the abundances of taxa relative to each other or their total—new studies are increasingly employing strategies to enable the analysis of cell density or other measures of "absolute abundance."
These relative and absolute abundances serve as the basis for a _differential-abundance (DA) analysis_, in which the change in abundance of a microbial taxon across samples or conditions is used to learn about the biology of the taxon and its impact on the host and other microbes as well as to detect predictive biomarkers of host and environmental health and disease.

<!-- P: Taxonomic bias -->
Although MGS-based DA analysis has been widely deployed and achieved many notable successes, it faces serious concerns over accuracy and reproducibility due to the inherent technical limitations of MGS measurements.
In particular, MGS measurements are _taxonomically biased_: Taxa vary dramatically (e.g. 10-1000X) in how efficiently they are measured—that is, converted from cells into taxonomically classified sequencing reads—by a given MGS protocol.
As a result, the abundance measurements obtained by MGS are inaccurate representations of the actual abundances and also tend to differ across protocols (@mclaren2019cons), studies, and even experimental batches (@yeh2018taxo). 
This bias arises from variation in how taxa respond to each step in an MGS protocol, from sample collection to bioinformatic classification.
Although often associated with variation in primer binding and amplification rates and marker-gene copy-number, large variation in DNA extraction efficiency and in the ability to correctly classify reads make taxonomic bias a feature of both shotgun and marker-gene measurements.
The error it causes have been found to in some cases to supersede sizable biological differences (e.g. @lozupone2013meta) and has plausibly caused replication failures for prominent findings such as the association of decreased Bacteroides and increased Firmicutes in stool with obesity (@finucane2014atax) and the association of certain taxa in the vagina of pregnant women with preterm birth (@callahan2017repl).

@@

The typical approach to countering taxonomic bias in DA analysis is to standardize the measurement protocol used within a given study.
In broad strokes, the thinking is that the measurements of samples measured by the same protocol will be affected by bias in the same way and so the inferred differences between samples (the focus of DA analysis) will be unaffected.
For example, if taxonomic bias consistently causes the measured proportion of a given species to be 10X too high, we can still accurately infer its fold changes across samples.
@kevorkian2018esti and @lloyd2020evid were the first (to our knowledge) to make this argument explicit mathematically; however, @mclaren2019cons used a theoretical model (supported by measurements of defined bacterial communities) to show that consistent taxonomic bias can lead to variable fold errors in measured proportions.
These varying errors can lead to spurious conclusions for how the proportion of a taxon varies across samples, even in the direction of change (for example, causing a taxon that decreases appear to increase).
Yet @mclaren2019cons also found that certain changes in relative abundance---in particular, fold changes in the ratios among species---are robust to the effects of bias.
The implications of these findings for changes in absolute abundance---which remain subject to taxonomic bias in the underlying MGS measurement---and for DA analysis across many species and many samples---as commonly done in microbiome association testing---have yet to be investigated.

@@

Here we use a combination of theoretical analysis, simulation, and re-analysis of published experiments to consider when and why taxonomic bias in MGS measurements leads to spurious results in DA analysis of relative and absolute abundance.
Our analysis clarifies how the folk wisdom that taxonomic bias does not affect the analysis of change across samples is only partially correct and can give a false sense of security in the accuracy of DA results.
Yet we also present several potential solutions---methods for quantifying, correcting, or otherwise accounting for the effect of taxonomic bias in DA analyses that can be deployed today with only modest changes to existing experimental and analytical workflows.
Over time, application of these methods to past and future experiments will provide crucial quantitative information about taxonomic bias and the conditions under which spurious results arise for various DA methodologies.
Collectively, these methods and insights may provide practical solutions to taxonomic bias in DA analysis and the confidence that is necessary to codify the statistical findings of microbiome studies into readily-translatable scientific knowledge.

>>>>

<!-- First paragraph: DA analysis -->
One of the most basic questions we can ask about microbial communities is: How do different microbial taxa vary in abundance—across space, time, and host or environmental conditions?
Advances in sequencing technology allow us to simultaneously measure the abundances of 100s to 1000s of species using marker-gene and shotgun metagenomic sequencing (jointly, MGS).
Although standard MGS measurements lose information about total microbial density—and so are typically used to analyze the abundances of taxa relative to each other or their total—new studies are increasingly employing strategies to enable the analysis of cell density or other measures of "absolute abundance."
These relative and absolute abundances serve as the basis for a _differential-abundance (DA) analysis_, in which the change in abundance of a microbial taxon across samples or conditions is used to learn about the biology of the taxon and its impact on the host and other microbes as well as to detect predictive biomarkers of host and environmental health and disease.

<!-- P: Taxonomic bias -->
Although MGS-based DA analysis has been widely deployed and achieved many notable successes, it faces serious concerns over accuracy and reproducibility due to the inherent technical limitations of MGS measurements.
In particular, MGS measurements are _taxonomically biased_: Taxa vary dramatically (e.g. 10-1000X) in how efficiently they are measured—that is, converted from cells into taxonomically classified sequencing reads—by a given MGS protocol (@mclaren2019cons).
As a result, the abundance measurements obtained by MGS are inaccurate representations of the actual abundances and also tend to differ across protocols, studies, and even experimental batches (@yeh2018taxo,@mclaren2019cons).
This bias arises from variation in how taxa respond to each step in an MGS protocol, from sample collection to bioinformatic classification.
Although often associated with variation in primer binding and amplification rates and marker-gene copy-number, large variation in DNA extraction efficiency and in the ability to correctly classify reads make taxonomic bias a feature of both shotgun and marker-gene measurements.
The error it causes have been found to in some cases to supersede sizable biological differences (e.g. @lozupone2013meta) and has plausibly caused replication failures for prominent findings such as the association of decreased Bacteroides and increased Firmicutes in stool with obesity (@finucane2014atax) and the association of certain taxa in the vagina of pregnant women with preterm birth (@callahan2017repl).

@@

The typical approach to countering taxonomic bias in DA analysis is to standardize the measurement protocol used within a given study.
In broad strokes, the thinking is that the measurements of samples measured by the same protocol will be affected by bias in the same way and so the inferred differences between samples (the focus of DA analysis) will be unaffected.
For example, if taxonomic bias consistently causes the measured proportion of a given species to be 10X too high, we can still accurately infer its fold changes across samples (@kevorkian2018esti,@lloyd2020evid).
However, @mclaren2019cons used a theoretical model (supported by measurements of defined bacterial communities) to show that consistent taxonomic bias can lead to variable fold errors in measured proportions.
These varying errors can lead to spurious conclusions for how the proportion of a taxon varies across samples, even in the direction of change (for example, causing a taxon that decreases appear to increase).
Yet @mclaren2019cons also found that the fold error in the ratios among species was constant, making fold changes in ratios robust to bias.
The implications of these findings for changes in absolute abundance---which remain subject to taxonomic bias in the underlying MGS measurement---and for DA analysis across many species and many samples---as commonly done in microbiome association testing---have yet to be investigated.

@@

Here we use a combination of theoretical analysis, simulation, and re-analysis of published experiments to consider when and why taxonomic bias in MGS measurements leads to spurious results in DA analysis of relative and absolute abundance.
Our analysis clarifies how the folk wisdom that taxonomic bias does not affect the analysis of change across samples is only partially correct and can give a false sense of security in the accuracy of DA results.
Yet we also present several potential solutions---methods for quantifying, correcting, or otherwise accounting for the effect of taxonomic bias in DA analyses that can be deployed today with only modest changes to existing experimental and analytical workflows.
Over time, application of these methods to past and future experiments will provide crucial quantitative information about the conditions under which taxonomic bias creates spurious results for various DA methodologies.
Collectively, these methods and insights may provide practical solutions to taxonomic bias in DA analysis and the confidence that is necessary to codify the statistical findings of microbiome studies into readily-translatable scientific knowledge.

### 2021-10-19 real-world-implications.Rmd
Make minor edits
Edits include

- use 'widehat' instead of 'hat' over words in equations
- use 'measurement' and 'measured' instead of 'estimate' and 'estimated'
  for individual samples and change between pairs of samples, reserving
  'estimate' for regression over multiple samples
- edits based on Hypothesis comments on version
  6932587bd1ea9c56638c0ebe3ff7885b79b84e8c

Partial but not full canceling of the effect of bias in proportion-based analyses, such that possible to have spurious results in principle; but what about in practice?
Predicting how frequently or which results are likely to be majorly affected is difficult because it depends on the joint variation in efficiencies of species and species abundance dynamics, both of which are poorly understood in most to all biological systems. 
However, through a combination of theoretical considerations, hypothetical scenarios, and real case studies we can begin to develop an understanding or the relevant considerations and build framework for improving our understanding as our knowledge of these dynamics improves.

@@

One approach to studying the mean efficiency empirically is to analyze data from studies for which control measurements allow direct measurement of species' efficiencies.
@brooks2015thet performed 16S rRNA gene sequencing of mock communities of seven bacterial species thought to play a critical role in the human vaginal microbiome with the MGS protocol developed for the Vaginal Human Microbiome Project (VaHMP). @mclaren2019cons later used this data to estimate the efficiencies of these seven species and develop a method to use these estimates to infer the true compositions of natural samples (see Section \@ref(#calibrate-compositions)).
SI Analysis (MOMS-PI) uses these estimated efficiencies to correct the compositions and explore the implied variation in mean efficiency in the VaHMP's Multi-Omic Microbiome Study: Pregnancy Initiative (MOMS-PI) @fettweis2019thev.
The vaginal microbiome during pregnancy is characterized by relatively low diversity, often with a single _Lactobacillus_ species often forming a majority of reads (considered to be the healthy state) but with occasional shifts to high proportions of other species, such as _Gardnerella vaginalis_, _Atopobium vaginae_, and _Streptococcus agalactiae_, which are thought to cause or indicate poor vaginal health and increased risk of preterm birth.
The analysis of @mclaren2019cons showed that the efficiencies of two common _Lactobacillus_ species, _L. crispatus_ and _L. iners_, were 20-30X greater than that of _G. vaginalis_, _A. vaginae_, and _S. agalactiae_.
Analysis of microbiome trajectories of a single patient across visits indicates that the mean efficiency typically decreases by a factor of 3-10X when the community shifts from a high to low-_Lactobacillus_ state.
(NOTE: This is a super preliminary claim based on looking at just one woman, and is just a placeholder)
Decreases in mean efficiency during transitions from _Lactobacillus_ to _Gardnerella_ dominance can be expected to be even more extreme for commonly-used vaginal microbiome primers that fail to amplify _Gardnerella_.

@@

Are there general mechanisms by which we might expect associations to arise between mean efficiency and the covariates commonly of interest in microbiome studies?
One possibility is that the ecology and measurement efficiency of species may become correlated simply due to shared evolutionary history.
A species' efficiency is heavily influenced by genetically determined traits such as cell-wall structure, ribosomal-operon copy number, sequence in a primer-targeted region, GC content, and even whether the species is present in a given taxonomic database.
Shared evolutionary history among phylogenetically-close species is expected and has been observed to create positive associations in these traits among more closely related species, such that we should expect a significant degree of phylogenetic conservation in efficiency.
Meanwhile, the same shared evolutionary history creates phylogenetic conservation in ecological traits that drives species abundance dynamics.
In this way, we may expect evolution to lead to a situation in which groups of species with similar efficiencies have similar shifts in abundance between conditions, creating the potential for commensurate shifts in mean efficiency.
For example, a change in diet that favors Bacteroidetes species relative to Firmicutes based on the resulting gut conditions will also increase the relative abundance of easy-to-lyse species.

Associations between efficiency and abundance can also arise more directly when a single microbial trait affects both.
For example, microbes at the ocean floor are slowly buried, sinking into a low nutrient, low oxygen environment.
@lloyd2020evid estimate the log fold change in the estimated absolute abundance of various taxa with sediment depth (as a proxy for time) to determine which taxa are able to persist and even grow in this difficult environment.
It is plausible that microbes with tougher cell walls would tend to persist longer (alive or dead) in the sediment, while at the same time being more difficult to extract DNA from than microbes with weaker cell walls.
As the relative abundance of tougher species increases with depth, the mean extraction efficiency decreases.
This decrease would increase the inferred log fold changes and could lead to inferred growth of taxa that are actually just persisting or even slowly dying off.
Another example of a trait that might simultaneously affect efficiency and relative abundance is ribosomal copy number, which increases the measurement efficiency in ribosomal amplicon experiments and is also linked to differences in ecology and population dynamics among species.

<!-- P: relative magnitude of species and mean efficiency changes -->
The biological significance of the error in $\hat \beta$ depends on the relative magnitude of the LFC (or covariance with the covariate $x$) in the focal species vs in the mean efficiency.
It may generally be true that the largest species LFCs in a DA analysis, which are often of primary interest in a microbiome study, will tend to be larger than the LFC of the mean efficiency, at least in high-diversity settings.
(This seems intuitively likely and true in examples, but I haven't analyzed it much mathematically and don't have a firm handle on this question.)
Further investigation into this question may improve our confidence in the top hits identified in published DA analyses meeting certain conditions.
However, we must remain weary of cases (as in the vaginal-microbiome and plant-fungal-interactions examples above) in which a shift in a single species can drive large changes in mean efficiency and lead to significant errors for all other species.
The experimental and computational methods discussed in Section \@ref(solutions) can mitigate taxonomic bias in these clearly-problematic cases while also helping to amass empirical evidence on the conditions in which bias is unlikely to cause major inferential errors.

>>>>

Are biologically significant errors in DA analyses likely in practice?
We can begin to answer this question using a combination of theoretical analysis, consideration of hypothetical scenarios, and analysis of case studies for which some information about taxonomic bias is available.

@@

One approach to studying the mean efficiency empirically is to analyze data from studies for which control measurements allow direct measurement of species' efficiencies.
@brooks2015thet performed 16S rRNA gene sequencing of mock communities of seven bacterial species thought to play a critical role in the human vaginal microbiome with the MGS protocol developed for the Vaginal Human Microbiome Project (VaHMP). @mclaren2019cons later used this data to estimate the efficiencies of these seven species and develop a method to use these estimates to infer the true compositions of natural samples (see Section \@ref(#calibrate-compositions)).
SI Analysis (MOMS-PI) uses these estimated efficiencies to correct the compositions and explore the implied variation in mean efficiency in the VaHMP's Multi-Omic Microbiome Study: Pregnancy Initiative (MOMS-PI) @fettweis2019thev.
The vaginal microbiome during pregnancy is characterized by relatively low diversity, with a single _Lactobacillus_ species often forming a majority of reads (considered to be the healthy state) but with occasional shifts to high proportions of other species, such as _Gardnerella vaginalis_, _Atopobium vaginae_, and _Streptococcus agalactiae_, which are thought to cause or indicate poor vaginal health and increased risk of preterm birth.
The analysis of @mclaren2019cons showed that the efficiencies of two common _Lactobacillus_ species, _L. crispatus_ and _L. iners_, were 20-30X greater than that of _G. vaginalis_, _A. vaginae_, and _S. agalactiae_.
Analysis of microbiome trajectories of a single patient across visits indicates that the mean efficiency typically decreases by a factor of 3-10X when the community shifts from a high to low-_Lactobacillus_ state.
(NOTE: This is a super preliminary claim based on looking at just one woman, and is just a placeholder)
Decreases in mean efficiency during transitions from _Lactobacillus_ to _Gardnerella_ dominance can be expected to be even more extreme for commonly-used vaginal microbiome primers that fail to amplify _Gardnerella_.

@@

Are there general mechanisms by which we might expect associations to arise between mean efficiency and the covariates commonly of interest in microbiome studies?
One possibility is that the ecology and measurement efficiency of species may become correlated simply due to shared evolutionary history.
A species' efficiency is heavily influenced by genetically determined traits such as cell-wall structure, ribosomal-operon copy number, sequence in a primer-targeted region, GC content, and even whether the species is present in a given taxonomic database.
Shared evolutionary history creates positive trait associations among more closely related species, such that we should expect a significant degree of phylogenetic conservation in efficiency.
Meanwhile, the same shared evolutionary history creates phylogenetic conservation in ecological traits that drives species abundance dynamics.
In this way, we may expect evolution to lead to a situation in which groups of species with similar efficiencies have similar shifts in abundance between conditions, creating the potential for commensurate shifts in mean efficiency.
For example, a change in diet that favors Bacteroidetes species relative to Firmicutes based on the resulting gut conditions will also increase the relative abundance of easy-to-lyse species.

Associations between efficiency and abundance can also arise more directly when a single microbial trait affects both.
For example, microbes at the ocean floor are slowly buried, sinking into a low nutrient, low oxygen environment.
@lloyd2020evid estimate the log fold change in the estimated absolute abundance of various taxa with sediment depth (as a proxy for time) to determine which taxa are able to persist and even grow in this difficult environment.
It is plausible that microbes with tougher cell walls would tend to persist longer (alive or dead) in the sediment, while at the same time being more difficult to extract DNA from than microbes with weaker cell walls.
As the relative abundance of tougher species increases with depth, the mean extraction efficiency might therefore decrease.
This decrease would increase the inferred log fold changes and could lead to inferred growth of taxa that are actually just persisting or even slowly dying off.
Another example of a trait that might simultaneously affect efficiency and relative abundance is ribosomal copy number, which increases the measurement efficiency in ribosomal amplicon experiments and is also linked to differences in ecology and population dynamics among species.

<!-- P: relative magnitude of species and mean efficiency changes -->
The biological significance of the error in $\hat \beta$ depends on the relative magnitude of the LFC (or covariance with the covariate $x$) in the focal species vs in the mean efficiency.
It may generally be true that the largest species LFCs in a DA analysis, which are often of primary interest in a microbiome study, will tend to be larger than the LFC of the mean efficiency, at least in high-diversity settings.
_(This seems intuitively likely and true in examples, but I haven't analyzed it much mathematically and don't have a firm handle on this question.)_
Further investigation into this question may improve our confidence in the top hits identified in published DA analyses meeting certain conditions.
However, we must remain weary of cases (as in the vaginal-microbiome and plant-fungal-interactions examples above) in which a shift in a single species can drive large changes in mean efficiency and lead to significant errors for other species.
The experimental and computational methods discussed in Section \@ref(solutions) can mitigate taxonomic bias in these clearly-problematic cases while also helping to amass empirical evidence on the conditions in which bias is unlikely to cause major inferential errors.
